---
title: "spotify_analysis_BScMathematics_s5513286"
output: html_document
date: "2026-04-15"
---

Data summary and variable normalization
```{r setup, include=FALSE}
library(dplyr)
library(moments)

#Import dataset
data_songs <- FINAL_

#Convert mode to a categorical variable, with values 0 or 1
data_songs$mode <- as.factor(data_songs$mode)


# Check skewness of the response variable
skewness(data_songs$popularity_beta)

# Summary of data 
data_songs  %>%
  group_by(Genre)  %>%
  summarise(
    n = n(),
    mean_popularity = mean(popularity_beta),
    sd_popularity = sd(popularity_beta),
    min = min(popularity_beta),
    max = max(popularity_beta)
  )

#Transform mode into categorical variable with values 0 and 1
data_songs$mode <- as.factor(data_songs$mode)

#Loudness z-score normalization
mu_1 <- mean(data_songs$loudness, na.rm = TRUE)
sigma_1 <- sd(data_songs$loudness, na.rm = TRUE)
loudness_z <- (data_songs$loudness - mu_1) / sigma_1


#Tempo z-score normalization
mu_2 <- mean(data_songs$tempo, na.rm = TRUE)
sigma_2 <- sd(data_songs$tempo, na.rm = TRUE)
tempo_z <- (data_songs$tempo - mu_2) / sigma_2

# Store normalized values in the dataset
data_songs$loudness_z <- loudness_z
data_songs$tempo_z <- tempo_z

```


Outliers for each predictor
```{r}
# IQR Outlier Detection Function
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  
  outliers <- x < lower_bound | x > upper_bound
  
  return(list(
    Q1 = Q1,
    Q3 = Q3,
    IQR = IQR_val,
    lower_bound = lower_bound,
    upper_bound = upper_bound,
    outliers = outliers,
    n_outliers = sum(outliers, na.rm = TRUE),
    n_total = sum(!is.na(x))
  ))
}

# Apply to variables
tempo_outlier_info <- detect_outliers(data_songs$tempo_z)
loudness_outlier_info <- detect_outliers(data_songs$loudness_z)
acousticness_outlier_info <- detect_outliers(data_songs$acousticness)
speechiness_outlier_info <- detect_outliers(data_songs$speechiness)
danceability_outlier_info <- detect_outliers(data_songs$danceability)
valence_outlier_info <- detect_outliers(data_songs$valence)
liveness_outlier_info <- detect_outliers(data_songs$liveness)
energy_outlier_info <- detect_outliers(data_songs$energy)

# Percentage function
get_pct <- function(n_out, n_total) {
  round(100 * n_out / n_total, 2)
}

# Results with percentages
cat("Tempo outliers:", tempo_outlier_info$n_outliers,
    "(", get_pct(tempo_outlier_info$n_outliers, tempo_outlier_info$n_total), "%)\n")

cat("Loudness outliers:", loudness_outlier_info$n_outliers,
    "(", get_pct(loudness_outlier_info$n_outliers, loudness_outlier_info$n_total), "%)\n")

cat("Acousticness outliers:", acousticness_outlier_info$n_outliers,
    "(", get_pct(acousticness_outlier_info$n_outliers, acousticness_outlier_info$n_total), "%)\n")

cat("Speechiness outliers:", speechiness_outlier_info$n_outliers,
    "(", get_pct(speechiness_outlier_info$n_outliers, speechiness_outlier_info$n_total), "%)\n")

cat("Danceability outliers:", danceability_outlier_info$n_outliers,
    "(", get_pct(danceability_outlier_info$n_outliers, danceability_outlier_info$n_total), "%)\n")

cat("Liveness outliers:", liveness_outlier_info$n_outliers,
    "(", get_pct(liveness_outlier_info$n_outliers, liveness_outlier_info$n_total), "%)\n")

cat("Energy outliers:", energy_outlier_info$n_outliers,
    "(", get_pct(energy_outlier_info$n_outliers, energy_outlier_info$n_total), "%)\n")

cat("Valnce outliers:", valence_outlier_info$n_outliers,
    "(", get_pct(valence_outlier_info$n_outliers, valence_outlier_info$n_total), "%)\n")
```

ICC of the null model
```{r}
library(glmmTMB)
library(performance)

# Null model including only the genre
null_model <- glmmTMB(
  popularity_beta ~ 1 + (1 | Genre),
  data   = data_songs,
  family = beta_family()
)

#ICC of the null model
icc_null <- icc(null_model)
print(icc_null)

```

Building Full Beta GLM and Full Beta BGLMM, followed by empirical justification of including random effects
```{r}
library(glmmTMB)

# Beta Generalized Linear Model
glm_beta_fixed <- glmmTMB(
  popularity_beta ~
    danceability +
    energy       +
    acousticness +
    valence      +
    speechiness  +
    liveness    +
    mode+
    loudness_z+
    tempo_z,
  data   = data_songs,
  family = beta_family()
)

#Full Beta Generalized Linear Mixed Model
glmm_beta <- glmmTMB(
  popularity_beta ~
    danceability +
    energy       +
    acousticness +
    valence      +
    speechiness  +
    liveness    +
    mode+
    loudness_z+
    tempo_z+
    (1 | Genre),
  data   = data_songs,
  family = beta_family(link = "logit")

)
summary(glmm_beta)

#ICC of the Full BGLMM
icc_full <- icc(glmm_beta)
print(icc_full)

#Model comparison between Full Beta GLM (including only fixed effects) and Full Beta GLMM (including both fixed and random effects)
anova(glm_beta_fixed, glmm_beta)
```
Backward elimination using minimum AIC procedure
```{r}
#Backward elimination using minimum AIC procedure 

library(MASS)

drop1(glmm_beta, test = "Chisq")
```
```{r}
#Step 1
glmm_beta_energy <- glmmTMB(
  popularity_beta ~
    danceability +
    acousticness +
    valence      +
    speechiness  +
    liveness    +
    mode+
    loudness_z+
    tempo_z+
    (1 | Genre),   # random intercept: accounts for genre clustering
  data   = data_songs,
  family = beta_family(link = "logit")

)
drop1(glmm_beta_energy, test = "Chisq")
```

```{r}
#Step 2
glmm_beta_tempo <- glmmTMB(
  popularity_beta ~
    danceability +
    acousticness +
    speechiness  +
    liveness    +
    mode+
    valence+
    loudness_z+
    (1 | Genre),   # random intercept: accounts for genre clustering
  data   = data_songs,
  family = beta_family(link = "logit")

)
drop1(glmm_beta_tempo, test = "Chisq")
```


```{r}
#Step 3
glmm_beta_loudness <- glmmTMB(
  popularity_beta ~
    danceability +
    acousticness +
    speechiness  +
    mode+
    valence+
    liveness+
    (1 | Genre),   
  data   = data_songs,
  family = beta_family(link = "logit")

)
drop1(glmm_beta_loudness, test = "Chisq")
```


```{r}
# Reduced Beta GLMM
glmm_beta_mode <- glmmTMB(
  popularity_beta ~
    danceability +
    acousticness +
    speechiness  +
    liveness+
    valence+
    (1 | Genre),   # random intercept: accounts for genre clustering
  data   = data_songs,
  family = beta_family(link = "logit")

)
drop1(glmm_beta_mode, test = "Chisq")
summary(glmm_beta_mode)
```


Three different Random Slope BGLMM are built with random slope for acousticness, liveness and speechiness respectively, which are the variables indetified having the most diverse relationship with popularity across genres. Using LRT we determine that only random slope for speechiness is significant to model fit.
```{r}
library(performance)

# Random Slope BGLMM (random slope for acousticness)
slope_acoust <- glmmTMB(
  popularity_beta ~ danceability +
    acousticness +
    speechiness  +
    valence+
    liveness+
    danceability+
  (1 + acousticness|Genre), family=beta_family(), data=data_songs)

# Random Slope BGLMM (random slope for liveness)
slope_liveness <- glmmTMB(
  popularity_beta ~ danceability +
    acousticness +
    speechiness  +
    valence+
    liveness+
    danceability+
  (1 + liveness|Genre), family=beta_family(), data=data_songs)

#Random Slope BGLMM (random slope for speechiness) - RS BGLMM
m_slope <- glmmTMB(
  popularity_beta ~ danceability +
    acousticness +
    speechiness  +
    valence+
    liveness+
    danceability+
  (1 + speechiness|Genre), family=beta_family(), data=data_songs)
summary(m_slope)

anova(slope_acoust, glmm_beta_mode)
anova(slope_liveness, glmm_beta_mode)
anova(m_slope, glmm_beta_mode)

```


```{r}
#Model comparisons 

#Full BGLMM against Reduced BGLMM
anova(glmm_beta, glmm_beta_red)

#Reduced BGLMM against RS BGLMM
anova(glmm_beta_red, m_slope, refit=FALSE)
```


```{r}
#Coefficient of Determination R^2
library(performance)

# Full model
r2_full <- r2(glmm_beta)
r2_full

# Reduced model
r2_reduced <- r2(glmm_beta_mode)
r2_reduced

# Random slope model
r2_slope <- r2(m_slope)
r2_slope
```

Cross-validation for the Full, Reduced, Random Slope BGLMM
```{r}
#m times repated k-fold stratified Cross-Validation for Full, Reduced, Random Slope BGLMM

library(caret) 
library(glmmTMB) 
library(lme4)

# Metric function 
compute_metrics <- function(y, y_hat) {
  
  valid <- is.finite(y) & is.finite(y_hat)
  
  if(sum(valid) == 0) {
    return(c(MSE=NA, RMSE=NA, MAE=NA))
  }
  
  y <- y[valid]
  y_hat <- y_hat[valid]
  
  mse  <- mean((y - y_hat)^2)
  rmse <- sqrt(mse)
  mae  <- mean(abs(y - y_hat))
  
  return(c(MSE = mse, RMSE = rmse, MAE = mae))
}

# Stratified CV
run_cv_stratified <- function(model_object, data, model_name) {
  
  all_results <- list()
  counter <- 1
  
  data$Genre <- as.factor(data$Genre)
  
  form <- formula(model_object)
  
  for (rep in 1:n_repeats) {
    set.seed(my_seed + rep)
    
    # Create stratified folds based on genre
    folds <- createFolds(data$Genre, k = K, list = TRUE)
    
    for (k in 1:K) {
      
      test_idx  <- folds[[k]]
      train_data <- data[-test_idx, ]
      test_data  <- data[test_idx, ]
      
      res <- tryCatch({
        
        if(inherits(model_object, "glmmTMB")) {
          model <- glmmTMB(
            form,
            data = train_data,
            family = model_object$modelInfo$family
          )
        } else if(inherits(model_object, "lmerMod")) {
          model <- lmer(form, data = train_data)
        }
        
        # Fixed-effects predictions
        y_hat <- predict(
          model,
          newdata = test_data,
          type = "response",
          re.form = NA
        )
        
        y <- test_data$popularity_beta
        
        metrics <- compute_metrics(y, y_hat)
        
        data.frame(
          Model = model_name,
          Repeat = rep,
          Fold = k,
          t(metrics)
        )
        
      }, error = function(e) {
        data.frame(
          Model = model_name,
          Repeat = rep,
          Fold = k,
          MSE=NA, RMSE=NA, MAE=NA
        )
      })
      
      all_results[[counter]] <- res
      counter <- counter + 1
    }
  }
  
  dplyr::bind_rows(all_results)
}
```


```{r}
# 10 times repeated stratified 10 fold CV for Full, Reduced, RS BGLMM

# Simulation Parameters
K <- 10
n_repeats <- 10
my_seed <- 42

# Run CV 
cv_full <- run_cv_stratified(
  glmm_beta,          
  data_songs,
  "Full"
)

cv_reduced <- run_cv_stratified(
  glmm_beta_mode,
  data_songs,
  "Reduced"
)

cv_slope <- run_cv_stratified(
  m_slope,
  data_songs,
  "Random_Slope"
)

# Combine and summarize
final_summary <- bind_rows(cv_full, cv_reduced, cv_slope) %>%
  group_by(Model) %>%
  summarise(
    across(
      c(MSE, RMSE, MAE),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        sd   = ~sd(.x, na.rm = TRUE)
      )
    ),
    .groups = "drop"
  )

print(final_summary)
```


Monte Carlo Simulation Study
```{r}
#Monte Carlo Simulation Study
library(glmmTMB) 
library(lme4)
library(performance)

# Define Gaussian GLMM
glmm_gauss<- glmmTMB(
  popularity_beta ~
    danceability +
    acousticness +
    valence +
    speechiness +
    liveness +
    (1 | Genre),
  data   = data_songs,
  family = gaussian(link = "identity")
)
 

#PHASE I

# 1. Set the baseline model 
B <- 100 
n_obs <- 2096
n_genres <- 8

# Fixed Effects
beta_int <- 0.817
beta_dance <- 0.141
beta_acoustic <- 0.213
beta_speech <- -0.553
beta_valence <- 0.115
beta_live <- -0.144

# Variance Components
sigma_genre <- 0.1635  
phi_true <- 13.5       

# 2. Simulation Loop
results <- vector("list", B)

for(i in 1:B) {

  

  # A. Generate predictors

  sim_data <- data.frame(
    Genre = factor(rep(1:n_genres, length.out = n_obs)),
    danceability = rnorm(n_obs),
    acousticness = rnorm(n_obs),
    speechiness  = rnorm(n_obs),
    valence      = rnorm(n_obs),
    liveness     = rnorm(n_obs)

  )

  # B. Random effects
  genre_effects <- rnorm(n_genres, 0, sigma_genre)
  sim_data$u_i <- genre_effects[as.numeric(sim_data$Genre)]

  

  # C. Linear predictor
  sim_data$eta <- with(sim_data,
    beta_int +
    beta_dance * danceability +
    beta_acoustic * acousticness +
    beta_speech * speechiness +
    beta_valence * valence +
    beta_live * liveness +
    u_i

  )

  sim_data$mu <- 1 / (1 + exp(-sim_data$eta))

  

  # D. Generate Beta response
  sim_data$y <- rbeta(
    n_obs,
    shape1 = sim_data$mu * phi_true,
    shape2 = (1 - sim_data$mu) * phi_true

  )

  

  # E. Fit models
  m_beta <- glmmTMB(

    y ~ danceability + acousticness + speechiness +valence + liveness + (1|Genre),
  data = sim_data,
  family = beta_family(link = "logit"))

  m_gauss <- lmer(y ~ danceability + acousticness + speechiness +valence + liveness + (1|Genre),
  data = sim_data)


  # F. Predictions
  preds_gauss <- predict(m_gauss)

  

  # G. Store results
  results[[i]] <- data.frame(
    est_beta_s = fixef(m_beta)$cond["speechiness"],
    est_gauss_s = fixef(m_gauss)["speechiness"],
    
    est_beta_a = fixef(m_beta)$cond["acousticness"],
    est_gauss_a = fixef(m_gauss)["acousticness"],
    
    est_beta_v = fixef(m_beta)$cond["valence"],
    est_gauss_v = fixef(m_gauss)["valence"],
    
    est_beta_d = fixef(m_beta)$cond["danceability"],
    est_gauss_d = fixef(m_gauss)["danceability"],
    
    est_beta_l = fixef(m_beta)$cond["liveness"],
    est_gauss_l = fixef(m_gauss)["liveness"],
    
    inadmissible = mean(preds_gauss < 0 | preds_gauss > 1),
    singular = isSingular(m_gauss)

  )

}

results_final <- do.call(rbind, results)

results_final <- bind_rows(results)

# PHASE I Summary
phase1_summary <- results_final %>%
  summarise(
    bias_beta_speech = mean(est_beta_s - beta_speech),
    bias_gauss_speech = mean(est_gauss_s - beta_speech),
    bias_beta_acoust = mean(est_beta_a - beta_acoustic),
    bias_gauss_acoust = mean(est_gauss_a - beta_acoustic),
    bias_beta_vale = mean(est_beta_v - beta_valence),
    bias_gauss_vale = mean(est_gauss_v - beta_valence),
    bias_beta_dance = mean(est_beta_d - beta_dance),
    bias_gauss_dance = mean(est_gauss_d - beta_dance),
    bias_beta_live = mean(est_beta_l - beta_live),
    bias_gauss_live = mean(est_gauss_l - beta_live),
    
    rmse_beta_speech = sqrt(mean((est_beta_s - beta_speech)^2)),
    rmse_gauss_speech = sqrt(mean((est_gauss_s - beta_speech)^2)),
    rmse_beta_acoust = sqrt(mean((est_beta_a - beta_acoustic)^2)),
    rmse_gauss_acoust = sqrt(mean((est_gauss_a - beta_acoustic)^2)),
    rmse_beta_vale = sqrt(mean((est_beta_v - beta_valence)^2)),
    rmse_gauss_vale = sqrt(mean((est_gauss_v - beta_valence)^2)),
    rmse_beta_dance = sqrt(mean((est_beta_d - beta_dance)^2)),
    rmse_gauss_dance = sqrt(mean((est_gauss_d - beta_dance)^2)),
    rmse_beta_live = sqrt(mean((est_beta_l - beta_live)^2)),
    rmse_gauss_live = sqrt(mean((est_gauss_l - beta_live)^2)),
    
    inadmissible_pct = mean(inadmissible) * 100,
    singularity_rate = mean(singular) * 100
  )

print(phase1_summary)
```


```{r}
library(glmmTMB)
library(lme4)

# PHASE II

control_params <- lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 1e4)
)

# Create grid
B <- 20
intercept_seq <- seq(-5, 5, by = 0.5)

failure_results <- data.frame()

for (b0 in intercept_seq) {

  inadmissible_vec <- numeric(B)
  singular_vec     <- numeric(B)
  mean_y_vec       <- numeric(B)

  for (b in 1:B) {

    # A. Generate data
    sim_data$eta <- b0 +
      beta_dance * sim_data$danceability +
      beta_acoustic * sim_data$acousticness +
      beta_speech * sim_data$speechiness +
      beta_valence * sim_data$valence +
      beta_live * sim_data$liveness +
      sim_data$u_i

    sim_data$mu <- 1 / (1 + exp(-sim_data$eta))

    sim_data$y <- rbeta(
      nrow(sim_data),
      sim_data$mu * phi_true,
      (1 - sim_data$mu) * phi_true
    )

    # Enforce (0,1) using beta transformation
    n <- nrow(sim_data)
    sim_data$y <- (sim_data$y * (n - 1) + 0.5) / n

    # B. Fit models
    m_beta_temp <- glmmTMB(
      y ~ danceability + acousticness + speechiness +
        valence + liveness + (1 | Genre),
      data = sim_data,
      family = beta_family(link = "logit")
    )

    m_gauss_temp <- lmer(
      y ~ danceability + acousticness + speechiness +
        valence + liveness + (1 | Genre),
      data = sim_data,
      control = control_params
    )

    # C. Predictions
    preds_gauss <- predict(m_gauss_temp)

    # D. Failure metrics
    out_low  <- sum(preds_gauss < 0)
    out_high <- sum(preds_gauss > 1)

    inadmissible_vec[b] <- (out_low + out_high) / length(preds_gauss) * 100
    singular_vec[b]     <- as.numeric(isSingular(m_gauss_temp))
    mean_y_vec[b]       <- mean(sim_data$y)
  }

  # E. Results
  failure_results <- rbind(failure_results, data.frame(
    intercept        = b0,
    mean_popularity  = mean(mean_y_vec),
    inadmissible_pct = mean(inadmissible_vec),
    singularity_rate = mean(singular_vec)
  ))
}

print(failure_results)
```


Cross-validation of the Gaussian GLMM
```{r}
library(dplyr)
library(caret)
library(glmmTMB)

#m times repeated k-fold stratified cross validation for Gaussian model
#Metric function
compute_metrics <- function(y, y_hat) {
  
  valid <- is.finite(y) & is.finite(y_hat)
  
  if(sum(valid) == 0) {
    return(c(MSE=NA, RMSE=NA, MAE=NA))
  }
  
  y <- y[valid]
  y_hat <- y_hat[valid]
  
  mse  <- mean((y - y_hat)^2)
  rmse <- sqrt(mse)
  mae  <- mean(abs(y - y_hat))
  
  return(c(MSE = mse, RMSE = rmse, MAE = mae))
}

#Gaussian Stratified CV
run_cv_gaussian <- function(model_object, data, model_name) {
  
  all_results <- list()
  counter <- 1
  
  data$Genre <- as.factor(data$Genre)
  form <- formula(model_object)
  
  for (rep in 1:n_repeats) {
    set.seed(my_seed + rep)
    
    folds <- createFolds(data$Genre, k = K, list = TRUE)
    
    for (k in 1:K) {
      
      test_idx  <- folds[[k]]
      train_data <- data[-test_idx, ]
      test_data  <- data[test_idx, ]
      
      res <- tryCatch({
        
        # Refit Gaussian GLMM
        model <- glmmTMB(
          form,
          data = train_data,
          family = gaussian()
        )
        
        # Predictions for fixed effects only
        y_hat <- predict(
          model,
          newdata = test_data,
          type = "response",
          re.form = NA   
        )
        
        y <- test_data$popularity_beta
        
        metrics <- compute_metrics(y, y_hat)
        
        data.frame(
          Model = model_name,
          Repeat = rep,
          Fold = k,
          t(metrics)
        )
        
      }, error = function(e) {
        data.frame(
          Model = model_name,
          Repeat = rep,
          Fold = k,
          MSE=NA, RMSE=NA, MAE=NA
        )
      })
      
      all_results[[counter]] <- res
      counter <- counter + 1
    }
  }
  
  bind_rows(all_results)
}
```


```{r}
# 10 times repeated 10 fold stratified cross validation for the Gaussian model
K <- 10
n_repeats <- 10
my_seed <- 42

cv_gaussian <- run_cv_gaussian(
  glmm_gauss,
  data_songs,
  "Gaussian GLMM"
)

#Results of CV for the Gaussian GLMM
gaussian_summary <- cv_gaussian %>%
  group_by(Model) %>%
  summarise(
    MSE_mean  = mean(MSE, na.rm = TRUE),
    MSE_sd    = sd(MSE, na.rm = TRUE),
    RMSE_mean = mean(RMSE, na.rm = TRUE),
    RMSE_sd   = sd(RMSE, na.rm = TRUE),
    MAE_mean  = mean(MAE, na.rm = TRUE),
    MAE_sd    = sd(MAE, na.rm = TRUE)
  )

print(gaussian_summary)
```


Rank-Based Chart Evaluation
```{r}
#Rank Chart Based Evaluation for the Reduced and RS BGLMMs

library(dplyr)
data_songs <- data_songs %>%
  mutate(
    pred_beta_reduced = predict(glmm_beta_mode,  type = "response"),
    pred_beta_slope   = predict(m_slope, type = "response")
  )

master_predictions <- data_songs %>%
  select(WeekID, `Week Position`,
         pred_beta_reduced,
         pred_beta_slope)

# A. Weekly rank correlations

weekly_rank_results <- master_predictions %>%
  filter(!is.na(`Week Position`)) %>%
  group_by(WeekID) %>%
  summarise(
    n_tracks = n(),

    # Spearman 
    spearman_beta_reduced = -cor(pred_beta_reduced, `Week Position`,
                                method = "spearman", use = "complete.obs"),
    spearman_beta_slope = -cor(pred_beta_slope, `Week Position`,
                              method = "spearman", use = "complete.obs"),

    # Kendall 
    kendall_beta_reduced = -cor(pred_beta_reduced, `Week Position`,
                               method = "kendall", use = "complete.obs"),
    kendall_beta_slope = -cor(pred_beta_slope, `Week Position`,
                             method = "kendall", use = "complete.obs"),

    .groups = "drop"
  )


# B. Filter weeks with less than 5 songs
weekly_rank_results_filtered <- weekly_rank_results %>%
  filter(n_tracks >= 5)


# C. Summary
global_rank_summary <- weekly_rank_results_filtered %>%
  summarise(
    # Means
    mean_spearman_beta_reduced = mean(spearman_beta_reduced, na.rm = TRUE),
    mean_spearman_beta_slope = mean(spearman_beta_slope, na.rm = TRUE),
    
    mean_kendall_beta_reduced = mean(kendall_beta_reduced, na.rm = TRUE),
    mean_kendall_beta_slope = mean(kendall_beta_slope, na.rm = TRUE),

    # Standard deviations 
    sd_spearman_beta_reduced = sd(spearman_beta_reduced, na.rm = TRUE),
    sd_spearman_beta_slope = sd(spearman_beta_slope, na.rm = TRUE),

    sd_kendall_beta_reduced = sd(kendall_beta_reduced, na.rm = TRUE),
    sd_kendall_beta_slope = sd(kendall_beta_slope, na.rm = TRUE),

  )

print(global_rank_summary)
```


Plots 

Figure 1
```{r}
#Distibution of Spotify popularity score Plot
ggplot(data_songs, aes(x = popularity_beta)) +
  geom_histogram(aes(y = ..density..),
                 bins  = 30,
                 fill  = "deepskyblue3",
                 color = "black",
                 alpha = 0.7) +
  geom_density(color    = "indianred3",
               linewidth = 0.7) +
  labs(
    title = "Distribution of Spotify Popularity Score",
    x     = "Popularity transformed (0,1)",
    y     = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title       = element_text(hjust = 0.5, face = "bold", size = 15),

    axis.text.x      = element_text(color = "black", size = 11, margin = margin(b = 15)),
    axis.text.y      = element_text(color = "black", size = 11),
    axis.title       = element_text(color = "black", face = "bold", size=13),

    panel.grid.major = element_line(color = "lightgrey"),
    panel.grid.minor = element_line(color = NA),

    panel.background = element_rect(fill = "#f7f7f7", color = NA),
    plot.background  = element_rect(fill = "#f7f7f7", color = NA)
  )

ggsave("popularity_plot.png", width = 8, height = 6, dpi = 300, bg = "#f7f7f7")

```


Figure 2
```{r}
#Popularity Distribution by Genre Plot

library(ggplot2)
ggplot(data_songs, aes(x = Genre, y = popularity_beta, fill = Genre)) +
  geom_boxplot(color = "black") +
  scale_fill_manual(values = c(
    "Pop"            = "aquamarine2",
    "Electronic"     = "coral",
    "Hip hop/ Rap"   = "chartreuse",
    "Country"        = "#9467bd",
    "Latin"          = "deeppink",
    "R&B"            = "darkorange",
    "Rock"           = "cadetblue",
    "Easy Listening" = "yellow1"
  )) +
  labs(
    title = "Popularity Distribution by Genre",
    x     = "Genre",
    y     = "Popularity transformed (0-1)"
  ) +
  theme_minimal() +
  theme(
    plot.title        = element_text(hjust = 0.5, face = "bold", size = 15),

    axis.text.x       = element_text(color = "black", size = 11, margin = margin(b = 15), angle=20),
    axis.text.y       = element_text(color = "black", size = 11),
    axis.title        = element_text(color = "black", face = "bold", size=13),

    legend.position   = "none",

    panel.grid.major  = element_line(color = "lightgrey"),
    panel.grid.minor  = element_line(color = NA),

    panel.background  = element_rect(fill = "#f7f7f7", color = NA),
    plot.background   = element_rect(fill = "#f7f7f7", color = NA)
  )

ggsave(
  filename = "genre_popularity_distributions.png",
  width    = 10,
  height   = 6,
  dpi      = 300,          
  bg       = "#f7f7f7"     
)
```


Figure 3 (only the scatter plot of speechiness against popularity is presented in the paper)
```{r}
library(ggplot2)

#Custom theme for scatterplots of audio features vs popularity across genres
theme_scatter<-theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    
    axis.text.x = element_text(color = "black", size = 11, margin = margin(b = 15)),
    axis.text.y = element_text(color = "black", size = 11),
    axis.title = element_text(color = "black", face = "bold", size=13),
    
    strip.text = element_text(face = "bold", size = 11, color = "black"),
    
    panel.grid.major = element_line(color = "lightgrey"),
    panel.grid.minor = element_line(color = NA),
    
    panel.background = element_rect(fill = "#f7f7f7", color = NA),
    plot.background = element_rect(fill = "#f7f7f7", color = NA),
    
    strip.background = element_rect(fill = "#e8e8e8", color = NA)
  )

#Scatterplots of speechiness vs popularity across genres

ggplot(data_songs, aes(x = speechiness, y = popularity_beta)) +
  geom_point(alpha = 0.3, size = 1, color = "deepskyblue3") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
  facet_wrap(~ Genre, ncol = 4) +
  labs(
    title = "Popularity vs Speechiness for each genre",
    x = "Speechiness",
    y = "Popularity transformed (0–1)"
  ) +
  theme_minimal() +
  theme_scatter
  

#ggsave("speechiness_popularity.png", width = 10, height = 6)

#Scatterplots of danceability vs popularity across genres
ggplot(data_songs, aes(x = danceability, y = popularity_beta)) +
  geom_point(alpha = 0.3, size = 1, color = "deepskyblue3") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
  facet_wrap(~ Genre, ncol = 4) +
  labs(
    title = "Popularity vs Danceability for each genre",
    x = "Danceability",
    y = "Popularity transformed (0–1)"
  ) +
  theme_minimal() +
  theme_scatter

#Scatterplots of acousticness vs popularity across genres
ggplot(data_songs, aes(x = acousticness, y = popularity_beta)) +
  geom_point(alpha = 0.3, size = 1, color = "deepskyblue3") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
  facet_wrap(~ Genre, ncol = 4) +
  labs(
    title = "Popularity vs Acousticness for each genre",
    x = "Acousticness",
    y = "Popularity transformed (0–1)"
  ) +
  theme_minimal() +
  theme_scatter

#Scatterplots of valence vs popularity across genres
ggplot(data_songs, aes(x = valence, y = popularity_beta)) +
  geom_point(alpha = 0.3, size = 1, color = "deepskyblue3") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
  facet_wrap(~ Genre, ncol = 4) +
  labs(
    title = "Popularity vs Valence for each genre",
    x = "Valence",
    y = "Popularity transformed (0–1)"
  ) +
  theme_minimal() +
  theme_scatter

#Scatterplots of liveness vs popularity across genres
ggplot(data_songs, aes(x = liveness, y = popularity_beta)) +
  geom_point(alpha = 0.3, size = 1, color = "deepskyblue3") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
  facet_wrap(~ Genre, ncol = 4) +
  labs(
    title = "Popularity vs Liveness for each genre",
    x = "Liveness",
    y = "Popularity transformed (0–1)"
  ) +
  theme_minimal() +
  theme_scatter

```


Figure 4
```{r}
# Decompose the variance into Fixed, Random, and Residual
r2_long <- data.frame(
  Model = c("Full", "Reduced", "Random Slope"),
  Fixed = c(r2_full$R2_marginal, r2_reduced$R2_marginal, r2_slope$R2_marginal),
  Random = c(r2_full$R2_conditional - r2_full$R2_marginal, 
             r2_reduced$R2_conditional - r2_reduced$R2_marginal, 
             r2_slope$R2_conditional - r2_slope$R2_marginal),
  Residual = c(1 - r2_full$R2_conditional, 
               1 - r2_reduced$R2_conditional, 
               1 - r2_slope$R2_conditional)
) %>%
  # Pivot to long format for ggplot
  pivot_longer(
    cols = c(Fixed, Random, Residual), 
    names_to = "Component", 
    values_to = "value"
  ) %>%
  mutate(label = percent(value, accuracy = 0.1))

# Variance Decomposiiton of Popularity Plot
ggplot(r2_long, aes(x = Model, y = value, fill = Component)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 4,
            fontface = "bold") + # Fixed 'face' to 'fontface'
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_fill_manual(values = c("Fixed" = "deepskyblue3", 
                               "Random" = "olivedrab3", 
                               "Residual" = "indianred1")) +
  labs(
    title = "Variance Decomposition of GLMMs",
    y = "Percentage",
    fill = "Component"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15), 
    axis.text.x = element_text(color = "black", size = 11, margin = margin(b = 15)),        
    axis.text.y = element_text(color = "black", size = 11),        
    axis.title = element_text(color = "black", face = "bold", size = 13),
    legend.background = element_rect(color = "black", linewidth = 0.5, fill = "#f7f7f7"),
    legend.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = NA),
    panel.grid.minor = element_line(color = NA),
    panel.background = element_rect(fill = "#f7f7f7", color = NA),
    plot.background = element_rect(fill = "#f7f7f7", color = NA)
  )


ggsave("variance_decomp.png", width = 8, height = 6, dpi = 300, bg = "#f7f7f7")
```


Figure 5
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

plot_data_all <- results %>%
  select(est_beta_s, est_gauss_s,
         est_beta_a, est_gauss_a) %>%
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Estimate") %>%
  mutate(
    Model = ifelse(grepl("beta", Variable), "Beta GLMM", "Gaussian GLMM"),
    Predictor = ifelse(grepl("_s", Variable), "Speechiness", "Acousticness")
  )

# True values dataframe
true_vals <- data.frame(
  Predictor = c("Speechiness", "Acousticness"),
  true = c(beta_speech, beta_acoustic)
)

#Distribution of Coefficient Estimates Across Simulations Plot
ggplot(plot_data_all, aes(x = Estimate, fill = Model, color = Model, linetype = Predictor)) +
  geom_density(alpha = 0.4) +
  geom_vline(data = true_vals,
             aes(xintercept = true, linetype = Predictor),
             size = 1,
             color = "black") +
  scale_fill_manual(values = c(
    "Beta GLMM"     = "deepskyblue3",
    "Gaussian GLMM" = "indianred1"
  )) +
  scale_color_manual(values = c(
    "Beta GLMM"     = "deepskyblue3",
    "Gaussian GLMM" = "indianred1"
  )) +
  scale_linetype_manual(values = c(
    "Speechiness"  = "solid",
    "Acousticness" = "dashed"
  )) +
  labs(
    title = "Distribution of Coefficient Estimates Across Simulations",
    x = "Estimated Coefficient",
    y = "Density",
    fill = "Model",
    color = "Model",
    linetype = "Predictor"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    axis.text.x = element_text(color = "black", size = 11, margin = margin(b = 15)),
    axis.text.y = element_text(color = "black", size = 11),
    axis.title = element_text(color = "black", face = "bold", size=13),
    legend.background = element_rect(color = "black", linewidth = 0.5, fill = "#f7f7f7"),
    legend.key = element_rect(fill = NA, color = NA),
    legend.text = element_text(size = 12),
    legend.title = element_text(face = "bold"),
    legend.box = "horizontal",          # <-- changed from "vertical"
    legend.spacing.x = unit(0.5, "cm"),
    legend.margin = margin(6, 6, 6, 6),
          legend.position = c(0.5, -0.24),
                              legend.direction="horizontal",
    # x=0 is far left, adjust y as needed
    plot.margin = margin(10, 10, 60, 10),
    panel.grid.major = element_line(color = "lightgrey"),
    panel.grid.minor = element_line(color = NA),
    panel.background = element_rect(fill = "#f7f7f7", color = NA),
    plot.background = element_rect(fill = "#f7f7f7", color = NA)
  )
ggsave("coefficient_distributions_combined.png", width = 9, height = 5)

```

Figure 6
```{r}
#Simulated Beta Response Distributions Plot

ggplot(demo_df, aes(x = y, fill = label, colour = label)) +
  geom_density(alpha = 0.20, linewidth = 0.85) +
  geom_vline(
    data      = vlines_df,
    aes(xintercept = m, colour = label),
    linetype  = "dashed",
    linewidth = 0.75
  ) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_fill_manual(values = c(
    "#e41a1c", "#ff7f00", "#4daf4a", "#377eb8", "#984ea3"
  )) +
  scale_colour_manual(values = c(
    "#e41a1c", "#ff7f00", "#4daf4a", "#377eb8", "#984ea3"
  )) +
  labs(
    title  = "Simulated Beta Response Distributions",
    x      = "Popularity score (y)",
    y      = "Density",
    fill   = NULL, 
    colour = NULL   
  ) +
  guides(
    fill   = guide_legend(ncol = 3),   
    colour = guide_legend(ncol = 3)
  ) +
  theme_minimal() +
  theme(
    plot.title        = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x       = element_text(color = "black", size = 10, margin = margin(b = 15)),
    axis.text.y       = element_text(color = "black", size = 10),
    axis.title        = element_text(color = "black", face = "bold"),
    legend.background = element_rect(color = "black", linewidth = 0.5, fill = "#f7f7f7"),
    legend.text       = element_text(size = 10),
    legend.title      = element_text(face = "bold"),
    legend.position   = "bottom",
    panel.grid.major  = element_line(color = "lightgrey"),
    panel.grid.minor  = element_line(color = NA),
    panel.background  = element_rect(fill = "#f7f7f7", color = NA),
    plot.background   = element_rect(fill = "#f7f7f7", color = NA)
  )

ggsave(
  filename = "simulated_distributions.png",
  width    = 8,
  height   = 5,
  dpi      = 300,          
  bg       = "#f7f7f7"     
)

```


Figure 7
```{r}
library(dplyr)
library(ggplot2)

#Failure criteria for Gaussian GLMM
failure_results <- failure_results %>%
  mutate(
    fail_flag = inadmissible_pct > 5 | singularity_rate > 10
  )

#Gaussian GLMM Percentage of Inadmissible Predictions and Singularity Rate

ggplot(failure_results, aes(x = intercept)) +
  geom_line(aes(y = inadmissible_pct, color = "Inadmissible predictions (%)"), size = 1.2) +
  geom_point(aes(y = inadmissible_pct, color = "Inadmissible predictions (%)"), size = 2) +
  
  # Singularity line + points
  geom_line(aes(y = singularity_rate, color = "Singularity Rate (%)"), size = 1.2) +
  geom_point(aes(y = singularity_rate, color = "Singularity Rate (%)"), size = 2) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-4, 4), linetype = "dotted", color = "black") +
  
  scale_color_manual(values = c(
    "Inadmissible predictions (%)" = "indianred1",
    "Singularity Rate (%)" = "deepskyblue3"
  )) +
  labs(
    title = "Gaussian GLMM Failure under Increasing Skewness",
    x = bquote(bold("Gobal fixed intercept β"[bold("0")])),
    y = "Percentage (%)",
    color = NULL
  ) +
  scale_y_continuous(
    limits = c(0, 26),
    breaks = seq(0, 26, by = 5)
  )+
  
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    
    axis.text.x = element_text(color = "black", size = 11, margin = margin(b = 15)),
    axis.text.y = element_text(color = "black", size = 11),
    axis.title = element_text(color = "black", face = "bold", size=13),
    
    legend.background = element_rect(color = "black", linewidth = 0.5, fill = "#f7f7f7"),
    legend.text = element_text(size = 12),
    legend.title = element_text(face = "bold"),
    legend.position   = "bottom",
    
    panel.grid.major = element_line(color = "lightgrey"),
    panel.grid.minor = element_line(color = NA),
    
    panel.background = element_rect(fill = "#f7f7f7", color = NA),
    plot.background = element_rect(fill = "#f7f7f7", color = NA)
  )

ggsave("glmm_failure.png", width = 8, height = 5)


```


Figure 8
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

weekly_long <- weekly_rank_results %>%
  filter(n_tracks >= 5) %>%
  pivot_longer(
    cols = c(spearman_beta_reduced, spearman_beta_slope),
    names_to = "Model",
    values_to = "Spearman_Rho"
  ) %>%
  mutate(
    Model = recode(Model,
                   spearman_beta_reduced = "Reduced BGLMM",
                   spearman_beta_slope   = "RS BGLMM")
  )

ggplot(weekly_long, aes(x = Model, y = Spearman_Rho, fill = Model, color = Model)) +
  geom_violin(alpha = 0.3,
              draw_quantiles = c(0.25, 0.75),
              linetype = "dotted") +
  geom_boxplot(width = 0.15,
               fill = "white",
               outlier.alpha = 0.4,
               outlier.size = 1) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 23,
               size = 3,
               fill = "black") +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             color = "grey50") +
  scale_fill_manual(values = c(
    "Reduced BGLMM" = "deepskyblue3",
    "RS BGLMM"      = "indianred1"
  )) +
  scale_color_manual(values = c(
    "Reduced BGLMM" = "deepskyblue3",
    "RS BGLMM"      = "indianred1"
  )) +
  labs(
    x     = "Model Specification",
    y     = expression(paste("Spearman's ", hat(rho)[t])),
    title = "Distribution of Weekly Rank Correlations"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),

    axis.text.x  = element_text(color = "black", size = 10, margin = margin(b = 15)),
    axis.text.y  = element_text(color = "black", size = 10),
    axis.title   = element_text(color = "black", face = "bold"),

    legend.position = "none",

    
    panel.grid.major = element_line(color = "lightgrey"),
    panel.grid.minor = element_line(color = NA),

    panel.background = element_rect(fill = "#f7f7f7", color = NA),
    plot.background  = element_rect(fill = "#f7f7f7", color = NA)
  )

ggsave("weekly_rank_correlations.png", width = 7, height = 5)
```


Figure 9
```{r}
#Fitted vs Residual Plots

#Full
ggplot(data.frame(
  fitted = fitted(glmm_beta),
  residuals = residuals(glmm_beta)
), aes(x = fitted, y = residuals)) +
  
  geom_point(alpha = 19, color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  
  labs(
    title = "Fitted vs Residuals (Full BGLMM)",
    x = "Fitted values",
    y = "Residuals"
  ) +
  
  my_theme

#Gaussian
ggplot(data.frame(
  fitted = fitted(glmm_gauss),
  residuals = residuals(glmm_gauss)
), aes(x = fitted, y = residuals)) +
  
  geom_point(alpha = 19, color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  
  labs(
    title = "Fitted vs Residuals (Gaussian GLMM)",
    x = "Fitted values",
    y = "Residuals"
  ) +
  
  my_theme
ggsave("resgauss.png", width = 8, height = 6, dpi = 300, bg = "#f7f7f7")

#Reduced
ggplot(data.frame(
  fitted = fitted(glmm_beta_mode),
  residuals = residuals(glmm_beta_mode)
), aes(x = fitted, y = residuals)) +
  
  geom_point(alpha = 19, color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  
  labs(
    title = "Fitted vs Residuals (Reduced BGLMM)",
    x = "Fitted values",
    y = "Residuals"
  ) +
  
  my_theme


#Random Slope
ggplot(data.frame(
  fitted = fitted(m_slope),
  residuals = residuals(m_slope)
), aes(x = fitted, y = residuals)) +
  
  geom_point(alpha = 19, color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  
  labs(
    title = "Fitted vs Residuals (Random Slope BGLMM)",
    x = "Fitted values",
    y = "Residuals"
  ) +
  
  my_theme
```


Figures 10, 11
```{r}
library(ggplot2)
library(patchwork)

#Scatter plots of each audio feature against popularity score

# Custom theme
my_theme <- theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),

    axis.text.x  = element_text(color = "black", size = 11, margin = margin(b = 15)),
    axis.text.y  = element_text(color = "black", size = 11),
    axis.title   = element_text(color = "black", size = 13),

    legend.background = element_rect(color = "black", linewidth = 0.5, fill = "#f7f7f7"),
    legend.key        = element_rect(fill = NA, color = NA),
    legend.text       = element_text(size = 10),
    legend.title      = element_text(face = "bold"),

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    panel.background = element_rect(fill = "#f7f7f7", color = NA),
    plot.background  = element_rect(fill = "#f7f7f7", color = NA)
  )


p1 <- ggplot(data_songs, aes(x = acousticness, y = popularity_beta)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
  labs(
    x = "Acousticness",
    y = "Popularity (0-1)"
  ) +
  my_theme

p2 <- ggplot(data_songs, aes(x = danceability, y = popularity_beta)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
  labs(
    x = "Danceability",
    y = "Popularity (0-1)"
  )+
  my_theme

p3 <- ggplot(data_songs, aes(x = energy, y = popularity_beta)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE)+
  labs(
    x = "Energy",
    y = "Popularity (0-1)"
  )+
  my_theme

p4 <- ggplot(data_songs, aes(x = liveness, y = popularity_beta)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
  labs(
    x = "Liveness",
    y = "Popularity (0-1)"
  )+
  my_theme

p5 <- ggplot(data_songs, aes(x = speechiness, y = popularity_beta)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
   labs(
    x = "Speechiness",
    y = "Popularity (0-1)"
  )+
  my_theme

p6 <- ggplot(data_songs, aes(x = valence, y = popularity_beta)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
   labs(
    x = "Valence",
    y = "Popularity (0-1)"
  )+
  my_theme

p7 <- ggplot(data_songs, aes(x = loudness_z, y = popularity_beta)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
   labs(
    x = "Loudness",
    y = "Popularity (0-1)"
  )+
  my_theme

p8 <- ggplot(data_songs, aes(x = tempo_z, y = popularity_beta)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(method = "lm", color = "indianred1", se = FALSE) +
   labs(
    x = "Tempo",
    y = "Popularity (0-1)"
  )+
  my_theme

((p1 + p2) / (p3 + p4))  +
  plot_annotation(
    title = "Relationship between Audio Features and Popularity",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 15)
    )
  )
ggsave(
  filename = "scatter1.png",
  width    = 8,
  height   = 6,
  dpi      = 300,          
  bg       = "#f7f7f7"     
)
  
 ((p5 + p6) / (p7 + p8)) +
  plot_annotation(
    title = "Relationship between Audio Features and Popularity",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 15)
    )
  )

ggsave(
  filename = "scatter2.png",
  width    = 8,
  height   = 6,
  dpi      = 300,          
  bg       = "#f7f7f7"     
)

```

Plots not presented on the paper, but are used complementary to the analysis.

The following two plots are used just to understand intuitively the Random Intercept and Random Slope models.
```{r}
library(ggeffects)
library(ggplot2)

# Generate predictions for the Full BGLMM

pred_intercept_speech <- ggpredict(glmm_beta, terms = c("speechiness", "Genre"), type = "random")

# Random Intercept Model Plot
plot(pred_intercept_speech) +
  scale_color_manual(values = c(
    "Pop"            = "aquamarine2",
    "Electronic"     = "coral",
    "Hip hop/ Rap"   = "chartreuse",
    "Country"        = "#9467bd",
    "Latin"          = "deeppink",
    "R&B"            = "darkorange",
    "Rock"           = "cadetblue",
    "Easy Listening" = "yellow1"
  )) +
  scale_fill_manual(values = c(
    "Pop"            = "aquamarine2",
    "Electronic"     = "coral",
    "Hip hop/ Rap"   = "chartreuse",
    "Country"        = "#9467bd",
    "Latin"          = "deeppink",
    "R&B"            = "darkorange",
    "Rock"           = "cadetblue",
    "Easy Listening" = "yellow1"
  )) +
  guides(color = guide_legend(override.aes = list(fill = NA))) +
  labs(
    title  = "Random Intercept Model",
    x      = "Speechiness",
    y      = "Predicted Popularity",
    colour = "Genre"
  ) +
  theme_minimal() +
 my_theme

#ggsave("random_intercept_model.png", width = 8, height = 5)


```


```{r}

# Predictions for RS BGLMM
pred_slope_speech <- ggpredict(m_slope, terms = c("speechiness", "Genre"), type = "random")

#Random Slope Model Predictions for speechiness
library(ggplot2)

plot(pred_slope_speech) +
  scale_color_manual(values = c(
    "Pop"            = "aquamarine2",
    "Electronic"     = "coral",
    "Hip hop/ Rap"   = "chartreuse",
    "Country"        = "#9467bd",
    "Latin"          = "deeppink",
    "R&B"            = "darkorange",
    "Rock"           = "cadetblue",
    "Easy Listening" = "yellow1"
  )) +
  scale_fill_manual(values = c(
    "Pop"            = "aquamarine2",
    "Electronic"     = "coral",
    "Hip hop/ Rap"   = "chartreuse",
    "Country"        = "#9467bd",
    "Latin"          = "deeppink",
    "R&B"            = "darkorange",
    "Rock"           = "cadetblue",
    "Easy Listening" = "yellow1"
  )) +
  labs(
    title = "Random Slope Model",
    x = "Speechiness",
    y = "Predicted Popularity"
  ) +
  guides(color = guide_legend(override.aes = list(fill = NA)))+
  theme_minimal() +
  my_theme

ggsave("random_slope.png", width = 8, height = 6, dpi = 300, bg = "#f7f7f7")
```


Plots just to understand the predictive ability of the models.
```{r}
library(lattice)
library(ggplot2)

#Predicted values of the three models
pred_full <- predict(glmm_beta, type = "response")
pred_red <- predict(glmm_beta_mode, type = "response")
pred_slope <- predict(m_slope, type = "response")
pred_gauss <- predict(glmm_gauss, type = "response")

#Predicted vs Observed Plots

#Full
ggplot(data.frame(pred = pred_full, obs = data_songs$popularity_beta),
       aes(x = pred, y = obs)) +
  
  geom_point(color = "cornflowerblue", size = 2) +
  
  geom_abline(intercept = 0, slope = 1, color = "black", linewidth = 1.2) +
  
  labs(
    title = "Predicted vs Observed Popularity for Full_BGLMM",
    x = "Predicted values",
    y = "Observed values"
  ) +
  theme_minimal()+
  my_theme

#Reduced
ggplot(data.frame(pred = pred_red, obs = data_songs$popularity_beta),
       aes(x = pred, y = obs)) +
  
  geom_point(color = "cornflowerblue", size = 2) +
  
  geom_abline(intercept = 0, slope = 1, color = "black", linewidth = 1.2) +
  
  labs(
    title = "Predicted vs Observed Popularity for Reduced_BGLMM",
    x = "Predicted values",
    y = "Observed values"
  ) +
  theme_minimal()+
  my_theme

#Random Slope
ggplot(data.frame(pred = pred_slope, obs = data_songs$popularity_beta),
       aes(x = pred, y = obs)) +
  
  geom_point(color = "cornflowerblue", size = 2) +
  
  geom_abline(intercept = 0, slope = 1, color = "black", linewidth = 1.2) +
  
  labs(
    title = "Predicted vs Observed Popularity for RS_BGLMM",
    x = "Predicted values",
    y = "Observed values"
  ) +
  theme_minimal()+
  my_theme

#Gaussian GLMM
ggplot(data.frame(pred = pred_gauss, obs = data_songs$popularity_beta),
       aes(x = pred, y = obs)) +
  
  geom_point(color = "cornflowerblue", size = 2) +
  
  geom_abline(intercept = 0, slope = 1, color = "black", linewidth = 1.2) +
  
  labs(
    title = "Predicted vs Observed Popularity for Gaussian GLMM",
    x = "Predicted values",
    y = "Observed values"
  ) +
  theme_minimal()+
  my_theme

```

The following plots are used to check the normality of residuals and random effects.
```{r}
#Checking normality of residuals using QQ plots

#Full
res_full <- residuals(glmm_beta)

ggplot(data.frame(res = res_full), aes(sample = res)) +
  stat_qq(color = "cornflowerblue") +
  stat_qq_line(color = "black", linewidth = 1) +
  labs(
    title = "QQ Plot of Residuals (Full BGLMM)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  my_theme

res_red <- residuals(glmm_beta_mode)

ggplot(data.frame(res = res_red), aes(sample = res)) +
  stat_qq(color = "cornflowerblue") +
  stat_qq_line(color = "black", linewidth = 1) +
  labs(
    title = "QQ Plot of Residuals (Reduced BGLMM)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  my_theme

#Reduced
res_red <- residuals(glmm_beta_mode)

ggplot(data.frame(res = res_red), aes(sample = res)) +
  stat_qq(color = "cornflowerblue") +
  stat_qq_line(color = "black", linewidth = 1) +
  labs(
    title = "QQ Plot of Residuals (Reduced BGLMM)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  my_theme

#Random Slope
res_red <- residuals(m_slope)

ggplot(data.frame(res = res_red), aes(sample = res)) +
  stat_qq(color = "cornflowerblue") +
  stat_qq_line(color = "black", linewidth = 1) +
  labs(
    title = "QQ Plot of Residuals (Random Slope BGLMM)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  my_theme
```

```{r}
#Checking normality of random intercepts and random slopes using QQ plots

#Full
re_full <- ranef(glmm_beta)$cond$Genre[,1]

ggplot(data.frame(re = re_full), aes(sample = re)) +
  stat_qq(color = "cornflowerblue") +
  stat_qq_line(color = "black", linewidth = 1) +
  labs(
    title = "QQ Plot of Random Intercepts (Full BGLMM)",
    x = "Theoretical Quantiles",
    y = "Random Intercepts"
  ) +
  my_theme

#Reduced
re_red <- ranef(glmm_beta_mode)$cond$Genre[,1]

ggplot(data.frame(re = re_full), aes(sample = re)) +
  stat_qq(color = "cornflowerblue") +
  stat_qq_line(color = "black", linewidth = 1) +
  labs(
    title = "QQ Plot of Random Intercepts (Reduced BGLMM)",
    x = "Theoretical Quantiles",
    y = "Random Intercepts"
  ) +
  my_theme

#Random Slopes
re_slope <- ranef(m_slope)$cond$Genre
intercepts <- re_slope[, "(Intercept)"]

ggplot(data.frame(re = intercepts), aes(sample = re)) +
  stat_qq(color = "cornflowerblue") +
  stat_qq_line(color = "black", linewidth = 1) +
  labs(
    title = "QQ Plot of Random Intercepts (RS Model)",
    x = "Theoretical Quantiles",
    y = "Random Intercepts"
  ) +
  my_theme

slopes <- re_slope[, "speechiness"]

ggplot(data.frame(re = slopes), aes(sample = re)) +
  stat_qq(color = "cornflowerblue") +
  stat_qq_line(color = "black", linewidth = 1) +
  labs(
    title = "QQ Plot of Random Slopes (Speechiness)",
    x = "Theoretical Quantiles",
    y = "Random Slopes"
  ) +
  my_theme
```
